---
title: "NAEP  Math 8th Grade 2011"
author: "Daniel Carr"
date: "October 16, 2017"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 0. Setup

```{r}
library(tidyverse)
library(micromapST)
```


## 1. Introduction

The document NAEP Nations Report Card Data Table Access
describes obtaining the data from the NCES web site.  

I download csv file and renamed it "MathG8Y2011.csv".

To discuss an easy and common way to prepare a csv 
file for statistical graphics production I use Excel 
remove sigDiff and sigSymbol columns and two rows (for DoDEA and
the National Pubic) and saved this in the file 
"MathG8Y2011Fixed.csv".  Removing the column was not
necessary, but it is not used in produce a linked
micromap but micromapST will not recognized
the two rows when trying to plot US states. 

The hand editing approach is often easy and successful
in the short terms, but often to leads to gaps reproducible
research trail.  While it may take a little more work
There merit in use an R script to fix the data because this provides documentation of what was done, support
reproduce the result and make changes later on.

Below we read the file to create a data.frame and show
how to used this in produce an introductory link micromaps
with three dot plot columns as preliminary.

With the data.frame in hand there are two basic tasks, 
creating micromap column description data.frame
and providing arguments to the micromapST function to
produce the plot. 

## 1.1 Read a valid state name file and NAEP csv file.name

```{r}
stateNames <- read.csv(file = 'stateNames.csv',
  header = TRUE,as.is = TRUE)[,1]

toFix <- read.csv(file = "MathG8Y2011.csv",
  header = TRUE,as.is = TRUE)
   
head(toFix)
```
## use select to remove varaiables and filter to remove rows
```{r}
mG8Y2011 <- toFix %>% 
  select(-(SigDiff:SigSymbol)) %>%
  filter(Jurisdiction %in% stateNames)
head(mG8Y2011)
```

## 1.1b Read the fixed data

```{r}
mG8Y2011Fixed <- read.csv(file = "MathG8Y2011Fixed.csv",
  header = TRUE,as.is = TRUE)

# compare the resulting data frames
all.equal(mG8Y2011, mG8Y2011Fixed)
```


## 1.2 Create a micromap column description data.frame
```{r}
## Run

colDesc <- data.frame(
type = c('map','id','dot','dot'),
lab1 = c('','','Average Points','Average Price (x10 USD)'),
#lab2 = c('' ,'','Mean'),
#lab3 = c('','','Possible 80-100','Percent),
col1 = c(NA,NA,'points','price')
)

t(colDesc)
```

In the column specificaton above, the micromap is to have
5 types of column.  They are, in left to right order, a map
column, and id column and three dot plot columns.

Each column can have two label lines above it and one
label line below it.  

The lab1 argument is a character string vector with
a character string for each column. The map and id
columns have automatic labels.  We enter two null
string ('', '' ,) as place holders. We want the top
label in the 3rd column to be blank so include another null
string ('',) as a place holder. We want the top labels
in the 4th and 5th columns to be 'Above'.   

The lab2 argument is a character string vector with a
character string for each column.  Again we include
empty character string as place holders for the map
and id columns. This time we label the three last columns.

The lab3 argument ia character string vector with a
character string for each column. The   
dot column bottom labels include the units of measure.

The col1 argument specifies the column values in the data.frame
to locate dots respective columns of the micromap.

A second micromap example specifies dotse as column types. 
Each micromap dotse column requires two columns
from data data.frame, one for mean and one for the standard error. 
plot. Hence the column description will have a col2 line
to specify the "se" data columns.  

There are more panel description arguments that can be included
in the column description file when needed.

## 1.3 Use the micromapST function to produce plot. 

Below the first two arguments are the data data.frame
and the column description data.frame.  

The next two arguments specify column with the
state ids and the type of id. 

The plotNames='full' puts the full statenames in the micromap.
the sortVar = 'MN' uses the variable to sort,
and ascend=FALSE makes the order descending.  
The title argument is a character string of length one or two.  

```{r echo=FALSE, fig.width=7.5, fig.height=10}
#pdf(file= "mG8Y2011Dots.pdf" , width=7.5, height=10)

micromapST(usavgscore, colDesc,
  rowNamesCol = 'province', rowNames = 'full',
  plotNames= 'full',
  sortVar = 'points',ascend=FALSE,
  title = c('US Wines Data',
          'Average Points and Prices')) 

#dev.off()
```


## 2. Dot plots with 95% confidence interval based on standard errors

## 2.1 Change the column description data.frame

The glyph use 'dotse'.  We also specify
the location of the standard errors in the  
data data.frame using the col2 vector.

```{r}
## Run

colDescSe <- data.frame(
type = c('map','id','dotse','dotse','dotse'),
lab1 = c('','','','Above','Above'),
lab2 = c('' ,'','Mean','Basic','Proficient'),
lab3 = c('','','Possible 0-500','Percent','Percent'),
col1 = c(NA,NA,'MN','AB','AP'),
col2 = c(NA,NA,'MN_SE','AB_SE','AP_SE')
  )

t(colDescSe)
```


## 2.2 This time open a pdf file, produce plot, and close the file.  

```{r }
pdf(file = "mG8Y2011DotsSe.pdf" , width = 7.5, height = 10)

micromapST(mG8Y2011, colDescSe,
  rowNamesCol = 1,
  rowNames = "full",
  plotNames = "full",
  sortVar = "MN",ascend = FALSE,
  title = c("NAEP Math Grade 8 in 2011",
          "Average Scale Scores and Achievement Percents"))

dev.off()
```

## 2.3 Dot and confidence intervals with lower and upper bounds 

The this case the use the dotconf glyph. There col1, col2, and
col3 rows colDesc data.frame to specify the data.frame column
for the estimate, the lower bound and the upper bound respectively

## 3. Map types, arrow glyphs and reference values

In the colDesc data.frame we can specify
four different types of maps:
map, mapcum, maptail and mapmedian.

The "map" version shown does the least to
reveal spatial patterns but is likely the
easiest to understand at first encounter.

Cumulative highlighted maps, "mapcum" is natural version to
introduce next. We reading the maps top down, once
state polygon is highlighted using as saturated color
and black outline, will continue to popped into
the foreground using a light yellow fill with
a black outline.

The accumulation for foreground state polygons
as we read downward from the top of the plot.
provide 10 highlighted polygon contours that
reveal clusters and shapes. The pattern can
motivate hypothesis generation about the
phenomena the behind the patterns.


The maptail option accumulates states from the top down
to the middle and from the bottom up to the middle.
Top down accumulation pops more and more states into
the foreground. Maps can look very busy when
more than half of the states are in the foreground.  
While having both the top and bottom as starting points for
accumulation toward the middle requires the reader to shift
perspective, experienced viewer may appreciate the
simpler appearance for the bottom half of the plot.

In the top half of the map column, "mapmedian" maps
highlight all the states at or above the median of
the sorting variable. In the bottom half of the
map column in highlights all of the states below
the map column.  (This assumes a decrease value map
order.)


Arrows provide a good way to show two values and
their difference.  Here we use arrows to compare males and
females.  Often we use arrows in a temporal context
to convey before and after values.  The arrow ends
use  position along a scale to encode two variables
and length to encode there difference.
(Arrow head are associated with an optical illusion
sour our ability to compared lengths accurately
may be reduce.)

A drawback of using arrows to show differences is
that the plot resolution for showing the difference.
The range of the two variables being compare sets the
comparison scale. The lengths of the difference may be
very small on this scale.

Using the arrows from zero for the Male - Female
greatly increases the resolution in the example below albeit
at the cost not directly provide the context of the original
values.

Sometimes there is a relationship between the difference of
two paired values and their average.  Looking a ratio of these
two transformed values can be informative, but not
shown in this example. The calculation is evident in the
script below but I remove this from plotting script that follows.

## Run

mathG8Gender13$Zero <- rep(0,nrow(mathG8Gender13))
# not helpful
# vM <- stateDF$Male
# vF <- stateDF$Female
# mathG8Gender13$RelDif <- (vM-vF)/(vM+vF)/2

panelDesc <- data.frame(
type=c('mapcum','id','arrow','arrow'),
lab1=c('','','Female To','0 To'),
lab2=c('' ,'','Male','Male-Female'),
col1 = c(NA,NA,13,17),
col2 = c(NA,NA,10,16),
refVals=c(NA,NA,NA,0))

fName = "MathG8Gender2013 Arrows.pdf"
pdf(file=fName,width=7.5,height=10)
micromapST(mathG8Gender13,panelDesc,
  sortVar=16,ascend=FALSE,
  title=c("NAEP Math Grade 8 in 2013",
          "Average Scale Scores"))

dev.off()


6. Maptail and Dots with confidence Intervals=======

As indicated above maptail maps accumulate from the
the top down to the middle and from the bottom up
to the middle.

The following creates good resolution a
300 dots per in inch .png file.
Such files may be easier to include
in Microsoft documents than .pdf files.
If this isn't supported on your operating
systems you may switch back to using
a pdf file for output.

## Run

panelDesc <- data.frame(
type=c('maptail','id','dotconf','dotconf'),
lab1=c('','','Male Estimates ','Female Estimates'),
lab2=c('' ,'','and 95% CIs','and 95% CIs'),
col1 = c(NA,NA,10,13),
col2 = c(NA,NA,11,14),
col3 = c(NA,NA,12,15))

fName = "Math G8 Gender 2013 DotConf.png"
png(file=fName,width=7,height=10,
    units="in",res=300)
micromapST(mathG8Gender13, panelDesc,
  sortVar=13,ascend=FALSE,
  title=c("NAEP Math Grade 8 in 2013",
          "Average Scale Scores"))

dev.off()

## End


7. More information is available about micromapSt

Some is in R.

## Run

?micromapST

## End

The micromapST.pdf file is paper entitled

micromapST: Exploring and Communicating
Geospatial Patterns in U.S. State Data

It has several examples showning options.
The examples can be instructive
Table 1 on page 11 covers
all the major options now available.

It is difficult to design in such limited
space.  There are aspect ratio and overplotting
problems.  The use of color for linking discourages
using color for other tasks that really need color
for fast visual processing. For the most part I suggest
emphasizing LM plots more like those in the book.


